---
title: "Employee Attrition Prediction with kNN and SVM"
author: "Hugo Blanco Iraola"
date: today
date-format: long
execute:
  echo: true
  warning: false
  error: false
  eval: true
  include: true
format:
  html:
    self-contained: true
    theme: default
    lang: en-US
---

# Employee Attrition Prediction with kNN and SVM

### Feature Engineering, Class Imbalance and Supervised Classification with tidymodels

This project develops a complete supervised machine learning workflow for predicting **employee attrition** using the `attrition` dataset from the R package `modeldata`.

The analysis focuses on a realistic People Analytics setting where the main objective is not only to maximize overall accuracy, but also to detect employees at risk of leaving the organization.

The workflow includes:

- Exploratory Data Analysis (EDA)
- Data-quality checks
- Stratified train/test splitting
- Feature engineering with `recipes`
- Class balancing with SMOTE
- k-Nearest Neighbors (kNN) hyperparameter tuning
- Polynomial Support Vector Machine (SVM)
- Evaluation using accuracy, sensitivity and specificity
- Model comparison from a business-oriented perspective

## 1. Dataset and Problem Definition

The dataset describes employee attrition and was originally provided by IBM Watson Analytics Lab. It is available directly through the `modeldata` package in R.

- **Objective:** predict whether an employee will leave the company.
- **Problem type:** supervised binary classification.
- **Target variable:** `Attrition` (`Yes` / `No`).
- **Predictors:** 30 employee-related variables.
- **Observations:** 1,470 employees.

```{r}
#| label: data-and-packages

library(tidyverse)
library(tidymodels)
library(modeldata)
library(skimr)
library(themis)

data(attrition, package = "modeldata")

glimpse(attrition)

attrition %>%
  summarise(
    n = n(),
    n_vars = ncol(.),
    duplicated_rows = sum(duplicated(.)),
    missing_cells = sum(is.na(.))
  )

attrition %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "n_na") %>%
  arrange(desc(n_na)) %>%
  filter(n_na > 0)
```

## 2. Exploratory Data Analysis

The initial dataset contains **1,470 observations and 31 variables**, with no duplicated rows or missing values.

The target variable is strongly imbalanced: approximately **83.9%** of employees belong to the `No` class and **16.1%** to the `Yes` class. This motivates the use of resampling techniques during model training.

```{r}
#| label: eda

skimr::skim(attrition)

attrition %>%
  count(Attrition) %>%
  mutate(prop = n / sum(n))

attrition %>%
  count(Attrition) %>%
  ggplot(aes(x = Attrition, y = n)) +
  geom_col() +
  labs(
    title = "Attrition Target Distribution",
    x = "Attrition",
    y = "Number of Employees"
  )

vars_num_demo <- c("Age", "MonthlyIncome", "YearsAtCompany")

attrition %>%
  select(Attrition, all_of(vars_num_demo)) %>%
  pivot_longer(-Attrition, names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = Attrition, y = value)) +
  geom_boxplot(outlier.alpha = 0.3) +
  facet_wrap(~ variable, scales = "free_y") +
  labs(
    title = "Numerical Predictors vs Attrition",
    x = "Attrition",
    y = "Value"
  )

attrition %>%
  count(Department, Attrition) %>%
  ggplot(aes(x = Department, y = n, fill = Attrition)) +
  geom_col(position = "fill") +
  labs(
    title = "Attrition by Department",
    x = "Department",
    y = "Proportion"
  ) +
  coord_flip()
```

The exploratory analysis suggests that several numerical and categorical variables differ between employees who leave and those who stay, indicating potentially useful predictive signal.

## 3. Train/Test Split

A fixed random seed is used to ensure reproducibility. Since employee attrition (`Yes`) is the event of interest, the target factor is explicitly releveled so that `Yes` becomes the first level.

The dataset is then divided into **70% training** and **30% test** sets using stratification by `Attrition`, preserving the original class prevalence.

```{r}
#| label: split

set.seed(123456)

attrition <- attrition %>%
  mutate(Attrition = relevel(Attrition, ref = "Yes"))

levels(attrition$Attrition)

split_attrition <- initial_split(
  attrition,
  prop = 0.7,
  strata = Attrition
)

train_data <- training(split_attrition)
test_data <- testing(split_attrition)

bind_rows(
  attrition %>% mutate(dataset = "Original"),
  train_data %>% mutate(dataset = "Train"),
  test_data %>% mutate(dataset = "Test")
) %>%
  count(dataset, Attrition) %>%
  group_by(dataset) %>%
  mutate(prop = n / sum(n))
```

The prevalence of `Attrition = Yes` remains practically constant across the original, training and test sets, confirming that stratification was applied correctly.

## 4. Feature Engineering and Class Balancing

A `tidymodels` recipe is built using the training data only in order to avoid information leakage.

The preprocessing pipeline includes:

1. Removal of zero- and near-zero-variance predictors.
2. Median imputation for numeric predictors if missing values are present.
3. Grouping of rare categorical levels below 1% frequency.
4. Dummy encoding of nominal predictors.
5. A second near-zero-variance filter after dummy creation.
6. Standardization of numeric predictors.
7. Removal of highly correlated numeric predictors using an absolute correlation threshold of 0.9.
8. SMOTE oversampling of the minority class during training.

```{r}
#| label: preprocessing-recipe

recipe_attrition <- recipe(Attrition ~ ., data = train_data) %>%
  step_nzv(all_predictors()) %>%
  step_impute_median(all_numeric_predictors()) %>%
  step_other(all_nominal_predictors(), threshold = 0.01) %>%
  step_dummy(all_nominal_predictors(), one_hot = FALSE) %>%
  step_nzv(all_predictors()) %>%
  step_normalize(all_numeric_predictors()) %>%
  step_corr(all_numeric_predictors(), threshold = 0.9) %>%
  themis::step_smote(Attrition, over_ratio = 0.6)

recipe_attrition

prep_recipe <- prep(recipe_attrition)
juice(prep_recipe) %>%
  count(Attrition)
```

SMOTE is applied only during model training, preventing synthetic observations from leaking into the held-out test set.

## 5. k-Nearest Neighbors Classification

The first classifier is a **k-Nearest Neighbors (kNN)** model. The number of neighbors, `k`, is treated as a hyperparameter and optimized using **10-fold stratified cross-validation**.

ROC AUC is used as the primary tuning metric because the target is imbalanced and ranking ability is important for identifying employees at risk.

```{r}
#| label: knn

knn_spec <- nearest_neighbor(
  neighbors = tune(),
  weight_func = "rectangular",
  dist_power = 2
) %>%
  set_engine("kknn") %>%
  set_mode("classification")

knn_wf <- workflow() %>%
  add_recipe(recipe_attrition) %>%
  add_model(knn_spec)

set.seed(123456)
folds <- vfold_cv(train_data, v = 10, strata = Attrition)

k_grid <- tibble(
  neighbors = unique(round(seq(3, 50, length.out = 10)))
)

knn_res <- tune_grid(
  knn_wf,
  resamples = folds,
  grid = k_grid,
  metrics = metric_set(roc_auc)
)

knn_res %>%
  collect_metrics() %>%
  arrange(desc(mean))

best_k <- select_best(knn_res, metric = "roc_auc")
best_k

knn_final_wf <- finalize_workflow(knn_wf, best_k)
knn_fit <- fit(knn_final_wf, data = train_data)

knn_test_pred <- predict(knn_fit, new_data = test_data, type = "prob") %>%
  bind_cols(predict(knn_fit, new_data = test_data, type = "class")) %>%
  bind_cols(test_data %>% select(Attrition))

knn_test_pred %>%
  conf_mat(truth = Attrition, estimate = .pred_class)

knn_metric_set <- metric_set(accuracy, sens, spec)

knn_metric_set(
  knn_test_pred,
  truth = Attrition,
  estimate = .pred_class
)

knn_test_pred %>%
  roc_curve(truth = Attrition, .pred_Yes) %>%
  autoplot()
```

### 5.1 kNN Results

The best value found in the explored grid is **k = 50**, with a mean cross-validated ROC AUC of approximately **0.77**.

On the test set, the model achieves approximately **64.7% accuracy** and identifies **51 employees who actually leave**, although this comes at the cost of a relatively high number of false positives.

This behavior is consistent with a retention-oriented People Analytics use case: the model favors identifying employees at risk, even if doing so generates additional false alarms.

Because the selected value lies at the upper boundary of the tested grid, a wider search range for `k` would be a reasonable next experiment.

## 6. Polynomial SVM Classification

The second classifier is a non-linear **Support Vector Machine with a polynomial kernel**.

The model uses fixed hyperparameters:

- `cost = 1`
- `degree = 2`
- `scale_factor = 1`

It is trained using the same preprocessing recipe as the kNN model, ensuring a fair comparison.

```{r}
#| label: svm

svm_spec <- svm_poly(
  cost = 1,
  degree = 2,
  scale_factor = 1
) %>%
  set_engine("kernlab") %>%
  set_mode("classification")

svm_wf <- workflow() %>%
  add_recipe(recipe_attrition) %>%
  add_model(svm_spec)

svm_fit <- fit(svm_wf, data = train_data)

svm_test_pred <- predict(svm_fit, new_data = test_data, type = "class") %>%
  bind_cols(test_data %>% select(Attrition))

svm_metric_set <- metric_set(accuracy, sens, spec)

svm_metric_set(
  svm_test_pred,
  truth = Attrition,
  estimate = .pred_class
)

svm_test_pred %>%
  conf_mat(truth = Attrition, estimate = .pred_class)
```

### 6.1 SVM Results

The polynomial SVM achieves approximately **81.7% accuracy** on the test set.

Its confusion matrix contains approximately:

- True positives: 27
- False positives: 36
- False negatives: 45
- True negatives: 334

The model therefore has high specificity but relatively low sensitivity for `Attrition = Yes`. It is conservative: it produces fewer false alarms, but it fails to detect many real employee departures.

## 7. Model Comparison

The two models exhibit very different classification behavior.

```{r}
#| label: model-comparison

knn_metrics <- knn_metric_set(
  knn_test_pred,
  truth = Attrition,
  estimate = .pred_class
) %>%
  mutate(model = "kNN")

svm_metrics <- svm_metric_set(
  svm_test_pred,
  truth = Attrition,
  estimate = .pred_class
) %>%
  mutate(model = "SVM")

bind_rows(knn_metrics, svm_metrics) %>%
  select(model, .metric, .estimate) %>%
  arrange(.metric, desc(.estimate))
```

The SVM obtains substantially higher overall accuracy because it performs very well on the majority `No` class. However, the business objective of this problem is to identify employees at risk of leaving, making **sensitivity for the `Yes` class** particularly important.

The kNN model detects considerably more real attrition cases, while the SVM sacrifices sensitivity in exchange for higher specificity and fewer false positives.

This illustrates why model selection in an imbalanced business problem should not rely on accuracy alone. The preferred model depends on the operational cost of false negatives versus false positives.

## 8. Conclusions

This project demonstrates an end-to-end classification workflow for employee attrition using the `tidymodels` ecosystem in R.

The main findings are:

- The attrition target is strongly imbalanced, making class-aware evaluation essential.
- A reusable preprocessing recipe can combine data cleaning, categorical encoding, scaling, correlation filtering and SMOTE without leaking information into the test set.
- kNN tuning with cross-validation produces a model with stronger sensitivity toward employees who leave.
- The polynomial SVM achieves substantially higher overall accuracy and specificity, but misses more real attrition cases.
- Accuracy alone is insufficient for selecting a model in this People Analytics context.
- The final model choice should reflect the business cost of missed departures versus unnecessary retention interventions.

Overall, the project highlights the importance of aligning model evaluation metrics with the real objective of the prediction task rather than optimizing a single aggregate metric.
